---
title: "Summarize colocalization filters"
subtitle: "Compute the number of SNPs and genes in our test set after various levels of filtering"
output: 
  html_document:
    toc: true
    toc_depth: 2
author: "Mike Gloudemans"
date: "`r format(Sys.time(), '%d %B, %Y')`"
---

```{r include=FALSE}
require(reshape2)
require(ggplot2)
require(dplyr)
require(readr)
```

We select the traits we are interested in. We originally load
all traits but then subset down to these.

```{r include=FALSE}
kept_traits = c("/users/mgloud/projects/insulin_resistance/data/gwas/formatted/brunas_sumstats/GWAS_FastInsu_adjBMI_MAGIC_Europeans.txt.gz",
                "/users/mgloud/projects/insulin_resistance/data/gwas/formatted/brunas_sumstats/GWAS_FastGlu_MAGIC_Europeans.txt.gz",
                "/users/mgloud/projects/insulin_resistance/data/gwas/formatted/brunas_sumstats/GWAS_MAGIC_ISI_Model_2_AgeSexBMI.txt.txt.gz",
                "/users/mgloud/projects/insulin_resistance/data/gwas/formatted/brunas_sumstats/GWAS_BMI_GIANT_Europeans.txt.gz",
                "/users/mgloud/projects/insulin_resistance/data/gwas/formatted/brunas_sumstats/GWAS_WHRadjBMI_GIANT_Europeans.txt.gz",
                "/users/mgloud/projects/insulin_resistance/data/gwas/formatted/bonus_sumstats/GWAS_T2D_DIAGRAM_European.txt.gz",
                "/users/mgloud/projects/insulin_resistance/data/gwas/formatted/brunas_sumstats/GWAS_TG_GCLC_Mixed.txt.gz",
                "/users/mgloud/projects/insulin_resistance/data/gwas/formatted/brunas_sumstats/GWAS_HDL_GCLC_Mixed.txt.gz",
                "/users/mgloud/projects/insulin_resistance/data/gwas/formatted/bonus_sumstats/GWAS_MI_adjBMI.txt.gz")

# For including in RMarkdown

trait_list = sapply(kept_traits, function(x) {s=strsplit(x, "/"); return(s[[1]][length(s[[1]])])})
names(trait_list) = NULL
```

```{r}
print(trait_list)
```



```{r include=FALSE}
kept_sub = sapply(kept_traits, function(x)
	{
		all = strsplit(x, "/")[[1]]
		trait = gsub("\\.", "_", all[length(all)])
	})
```

# Pre-colocalization filtering

Count the total number of unique SNPs being tested that were the lead SNP for at least one GWAS.
Note: We filter out the SNPs that were selected on the basis of excluded GWAS, meaning the only
ones we have left will be significant in at least one of the GWAS we care about (that was an
inclusion criterion for this list).

### Get number of unique GWAS lead SNPs.

```{r}
all_gwas_snps = read.table("/users/mgloud/projects/insulin_resistance/output/snps_considered.txt", col.names=c("chr", "snp.pos", "pvalue", "trait", "gwas_file"))
all_gwas_snps = all_gwas_snps[all_gwas_snps$gwas_file %in% kept_traits,]
all_gwas_snps$ref_snp = paste(all_gwas_snps$chr, all_gwas_snps$snp.pos, sep="_")
length(unique(all_gwas_snps$ref_snp))
write.table(all_gwas_snps, file="/users/mgloud/projects/insulin_resistance/output/all_gwas_snps.txt", quote=FALSE, col.names=TRUE, row.names=FALSE, sep="\t")
```

### Count the number of SNPs considered for each GWAS.

```{r}
hits_per_trait = all_gwas_snps %>% group_by(trait) %>% summarize(snp_count = length(unique(ref_snp)))
hits_per_trait$trait = sapply(as.character(hits_per_trait$trait), function(x) {s=strsplit(x, "/"); return(s[[1]][length(s[[1]])])})
print(hits_per_trait)
```

```{r include=FALSE}
# Count the number of unique loci that we're testing
# Function inputs a vector of SNPs and clusters them into locis
# by distance
group_to_loci = function(x)
{
        ids = unique(as.character(x))
	ids = ids[order(ids)]
        chr = sapply(as.character(ids), function(x) {strsplit(x, "_")[[1]][1]})
        pos = as.numeric(sapply(as.character(ids), function(x) {strsplit(x, "_")[[1]][2]}))
        loc_nums = rep(0, length(ids))
        loc_nums[1] = 1
        for (i in 2:length(ids))
        {
                # Check if there's a SNP above in the list within 1 MB of this SNP]
                same = (chr[1:(i-1)] == chr[i]) & (abs(pos[1:(i-1)] - pos[i]) < 1000000)
                if (length(which(same)) != 0)
                {
                        loc_nums[i] = loc_nums[which(same)[1]]
                }
                else
                {
                        loc_nums[i] = max(loc_nums) + 1
                }
        }

	mapped_loci = sapply(x, function(j) 
	{
			loc_nums[which(ids == j)]
        })

	return(mapped_loci)
}
```

### How many individual loci are represented in our collection of SNP?

```{r}
all_gwas_snps$locus = group_to_loci(all_gwas_snps$ref_snp)
print(length(unique(all_gwas_snps$locus)))
```

### Count the total number of genes found before filtering by eQTL p-values.

NOTE: Again, to be in this list a SNP-gene pair must be significant at least in the GWAS,
so no SNPs that shouldn't be tested will slip through this step.


```{r}
all_snp_gene_pairs = read.table("/users/mgloud/projects/insulin_resistance/output/snp_gene_pairs_considered.txt", col.names=c("chr", "snp.pos", "pvalue", "trait", "gwas_file", "feature", "eqtl_file"))
all_snp_gene_pairs = all_snp_gene_pairs[all_snp_gene_pairs$gwas_file %in% kept_traits,]
all_snp_gene_pairs$ref_snp = paste(all_snp_gene_pairs$chr, all_snp_gene_pairs$snp.pos, sep="_")
all_snp_gene_pairs$locus = group_to_loci(all_snp_gene_pairs$ref_snp)
# Total number of unique genes to test
length(unique(all_snp_gene_pairs$feature))
# Number of unique SNP-gene pairs to test
length(unique(paste(all_snp_gene_pairs$ref_snp, all_snp_gene_pairs$feature, sep="_")))
# Number of unique locus-gene pairs to test
length(unique(paste(all_snp_gene_pairs$locus, all_snp_gene_pairs$feature, sep="_")))

write.table(all_snp_gene_pairs, file="/users/mgloud/projects/insulin_resistance/output/all_snp_gene_pairs_prefiltering.txt", quote=FALSE, col.names=TRUE, row.names=FALSE, sep="\t")
```

### Count the list of tests we actually tried to run (after eQTL thresholding).

This also tells us which ones we've lost.
Once more, after filtering, SNPs will only be in this list if they were significant in
at least one of the GWAS we care about.

```{r}
testable_snps = read.table("/users/mgloud/projects/insulin_resistance/output/snps_to_test.txt", header=TRUE)
testable_snps = testable_snps[testable_snps$gwas_file %in% kept_traits,]
testable_snps$ref_snp = paste(testable_snps$chr, testable_snps$snp_pos, sep="_")
testable_snps$locus = group_to_loci(testable_snps$ref_snp)
testable_snps$trait = sapply(as.character(testable_snps$trait), function(x) {s=strsplit(x, "/"); return(s[[1]][length(s[[1]])])})
# Number of testable SNPs after filtering with eQTLs:
print(length(unique(testable_snps$ref_snp)))
# Number of testable loci after filtering:
print(length(unique(testable_snps$locus)))
# Number of testable genes after filtering:
print(length(unique(testable_snps$gene)))
# Number of unique SNP-gene pairs to test after filtering:
length(unique(paste(testable_snps$ref_snp, testable_snps$feature, sep="_")))
# Number of unique locus-gene pairs to test after filtering
length(unique(paste(testable_snps$locus, testable_snps$feature, sep="_")))
```

### How many SNPs per trait are now remaining?

```{r}
remaining_hits_per_trait = testable_snps %>% group_by(trait) %>% summarize(snp_count = length(unique(ref_snp)))
#print(remaining_hits_per_trait$trait)
print(data.frame(remaining_hits_per_trait))
```

```{r include=FALSE}
# Figure out which SNPs we dropped from the first list by eQTL thresholding
dropped_snps = all_gwas_snps[!(all_gwas_snps$ref_snp %in% unique(testable_snps$ref_snp)),]

# How many SNPs dropped per trait after applying eQTL filters?
dropped_hits_per_trait = dropped_snps %>% group_by(trait) %>% summarize(snp_count = length(unique(ref_snp)))
print(dropped_hits_per_trait)
```

# Post-colocalization summarization and QC

(We load all the colocalization files, in code not shown in the HTML document.)

```{r include=FALSE}
# Load all files into R for analysis
folders = dir("/users/mgloud/projects/brain_gwas/output/insulin-full-analysis-rerun/")
tabs = list()
err_tab = list()
skip_tab = list()
i = 1
for (folder in folders)
{
	files = dir(paste0("/users/mgloud/projects/brain_gwas/output/insulin-full-analysis-rerun/", folder))
	error = files[grep("ERROR", files)]
	skip = files[grep("skip", files)]
	files = files[grep("clpp" ,files)]
	
	if (length(error) > 0)
	{
		err_tab[[i]] = read.table(paste("/users/mgloud/projects/brain_gwas/output/insulin-full-analysis-rerun", folder, error, sep="/"), header=FALSE, sep="\t", fill=TRUE, col.names = c("gwas_file", "eqtl_file", "snp.chrom", "snp.pos", "restrict_gene", "trait", "error", "error2"))
		err_tab[[i]]$error = paste(err_tab[[i]]$error, err_tab[[i]]$error2, sep="|")
	}
	if (length(skip) > 0)
	{
		skip_tab[[i]] = read.table(paste("/users/mgloud/projects/brain_gwas/output/insulin-full-analysis-rerun", folder, skip, sep="/"), header=FALSE, sep="\t", fill=TRUE, col.names = c("gwas_file", "eqtl_file", "snp.chrom", "snp.pos", "feature", "error", "gwas_data"))
	}

	for (file in files)
	{
		tabs[[i]] = read.table(paste("/users/mgloud/projects/brain_gwas/output/insulin-full-analysis-rerun", folder, file, sep="/"), header=TRUE)
		i = i + 1
	}
}

results = do.call(rbind, tabs)
errors = do.call(rbind, err_tab)
skips = do.call(rbind, skip_tab)

# A bit of stuff just for more understandable labels
results$ensembl = sapply(as.character(results$feature), function(x) {strsplit(x, "\\.")[[1]][1]})
results$gwas_trait = gsub(".prepared.txt.gz", "", results$gwas_trait)
results$gwas_trait = gsub("_Mixed", "", results$gwas_trait)
results$gwas_trait = gsub("_Europeans", "", results$gwas_trait)
results$gwas_trait = gsub("_European", "", results$gwas_trait)
results$gwas_trait = gsub("_AllSNPs", "", results$gwas_trait)
results$gwas_trait = gsub("_MAGIC", "", results$gwas_trait)
results$gwas_trait = gsub("_GCLC", "", results$gwas_trait)
results$gwas_trait = gsub("_DIAGRAM", "", results$gwas_trait)
results$gwas_trait = gsub("_GIANT", "", results$gwas_trait)
results$gwas_trait = gsub("_CARDIoGRAMplusC4D", "", results$gwas_trait)
results$gwas_trait = gsub(".txt", "", results$gwas_trait)
results$gwas_trait = gsub("/users/mgloud/projects/gwas/data/prepared/", "", results$gwas_trait)
results$gwas_trait = gsub("/users/mgloud/projects/gwas/data/GENESIS/formatted/", "", results$gwas_trait)
results$gwas_trait = gsub("GWAS_", "", results$gwas_trait)
results$gwas_trait = gsub(".txt.gz", "", results$gwas_trait)
results$gwas_trait = gsub(".gz", "", results$gwas_trait)
results$eqtl_file = gsub("_allpairs_txt_gz", "", results$eqtl_file)

# Filter down to the ones that are in kept traits list
results = results[results$base_gwas_file %in% kept_sub,]

results = results[results$ref_snp %in% unique(all_gwas_snps$ref_snp),]
dim(results)
```

### How many SNPs were dropped during the colocalization analysis?

```{r}
dropped = testable_snps[!(testable_snps$ref_snp %in% unique(results$ref_snp)),]
dim(dropped)[1]
```

### Which SNPs were tested even though they don't appear to pass COLOC threshold in any tissue? 

Why? (Since we apply a filter at the beginning, this suggests that
they've been dropped because their lead variant was not measured in the eQTL study.)

```{r include=FALSE}
best_pvals = results %>% group_by(ref_snp, feature) %>% summarize(best_gwas_pval = max(X.log_gwas_pval), best_eqtl_pval = max(X.log_eqtl_pval))

best_pvals$best_gwas = sapply(1:dim(best_pvals)[1], function(i)
       {
		gene = best_pvals$feature[i]
       		pval = best_pvals$best_gwas_pval[i]
	        snp = best_pvals$ref_snp[i]

		best_pval = paste(unique(results[results$ref_snp == snp & abs(results$X.log_gwas_pval - pval) <= 0.01 & results$feature == gene,]$gwas_trait), collapse="-")


		return(best_pval)
       })

best_pvals$best_eqtl = sapply(1:dim(best_pvals)[1], function(i)
       {
		gene = best_pvals$feature[i]
       		pval = best_pvals$best_eqtl_pval[i]
	        snp = best_pvals$ref_snp[i]

		best_pval = paste(unique(results[results$ref_snp == snp & abs(results$X.log_eqtl_pval - pval) <= 0.01 & results$feature == gene,]$eqtl_file), collapse="-")
		return(best_pval)
       })

# For later, we'll also want to know all traits in which the GWAS and the eQTL p-values were significant.
results$all_sig_gwas = sapply(1:dim(results)[1], function(i)
	{
		feature = results$feature[i]
		ref_snp = results$ref_snp[i]
		sub = results[(results$feature == feature) & (results$ref_snp == ref_snp),]

		gwas_sig = sub[(sub$X.log_gwas_pval >= -log10(5*10^(-8))) | ((sub$X.log_gwas_pval >= 6) & (grepl("MI_", sub$gwas_trait) | (grepl("ISI", sub$gwas_trait)))),]
		return(paste(unique(gwas_sig$gwas_trait), collapse=";"))
	}
)

results$all_sig_eqtl = sapply(1:dim(results)[1], function(i)
	{
		feature = results$feature[i]
		ref_snp = results$ref_snp[i]
		sub = results[(results$feature == feature) & (results$ref_snp == ref_snp),]

		eqtl_sig = sub[(sub$X.log_eqtl_pval >= 5),]
		return(paste(unique(eqtl_sig$eqtl_file), collapse=";"))
	}
)



```

### Get the list of SNPs that shouldn't have actually been tested 

For these, the lead eQTL or GWAS SNP
was dropped during the intersection phase.

```{r}
insignificant_eqtl = best_pvals[best_pvals$best_eqtl_pval < 5,]
# Significant eQTL was dropped
print(dim(insignificant_eqtl)[1])
print(insignificant_eqtl)

insignificant_gwas = best_pvals[best_pvals$best_gwas_pval < -log10(5*10^(-8)),]
# If the SNP was significant in MI or in ISI, then we allow it as long as it passes best_gwas_pval < 10^(-6).
insignificant_gwas$allowed = sapply(1:dim(insignificant_gwas)[1], function(i)
{
	if (insignificant_gwas$best_gwas_pval[i] < 6)
	{
		return(FALSE)
	}
	ref = insignificant_gwas$ref_snp[i]
	feature = insignificant_gwas$feature[i]
	return(sum(grepl("_ISI_|_MI", all_gwas_snps[all_gwas_snps$ref_snp == ref,]$trait)) > 0)
})
insignificant_gwas = insignificant_gwas[!insignificant_gwas$allowed,]

# Significant GWAS was dropped
print(dim(insignificant_gwas)[1])
print(insignificant_gwas)

# Note: If we want to follow up on a particular gene to see why it was dropped, we
# can do it like so:
# one_insig = results[(results$feature == "ENSG00000002919.10") & (results$ref_snp == "17_46292923"),]
```

### Filter our results table to exclude the SNPs that are no longer significant.

```{r}
dim(results)[1]
results = results[!(paste(results$ref_snp, results$feature, sep="_") %in% paste(insignificant_eqtl$ref_snp, insignificant_eqtl$feature, sep="_")),]
results = results[!(paste(results$ref_snp, results$feature, sep="_") %in% paste(insignificant_gwas$ref_snp, insignificant_gwas$feature, sep="_")),]
dim(results)[1]
```

```{r include=FALSE}
# Now that we have the final list of colocalization results,
# we can assign a unique locus number to each locus, grouping
# nearby SNPs.
results$locus = group_to_loci(results$ref_snp)
```

### Which SNPs were not tested for all trait-tissue combos? 

Why? (Maybe due to errors while running the pipeline -- see if these errors
are excusable.)

```{r}
tests_per_pair = results %>% group_by(ref_snp, feature) %>% summarize(tests = length(feature))
missing_pairs = tests_per_pair[tests_per_pair$tests < (length(unique(results$eqtl_file)) * length(unique(results$gwas_trait))),]
# Total number of snp-gene pairs tested:
dim(tests_per_pair)[1]
# Total number of snp-gene pairs missing at least one trait-tissue combo:
dim(missing_pairs)[1]
all_missing_tests = results[paste(results$ref_snp, results$feature, sep="_") %in% paste(missing_pairs$ref_snp, missing_pairs$feature, sep="_"),]
```

I manually inspected the tests that were dropped. Based on the number of tests missed
for each locus, it's clear that dropped SNPs either occurred because the gene was not
tested for eQTLs in every trait, or because the GWAS summary stats at that locus had no overlap with the
eQTL summary statistics for at least one of the GWAS analyses.

For a bit more detail...

### Why are tests missing for some SNP-gene pairs?

For snp-gene pairs that have missing tests, is it because there was an error thrown in the pipeline,
or because the variant was intentionally skipped due to non-overlap or being on the edge of the range?

```{r}
missing_pairs$error = paste(missing_pairs$ref_snp, missing_pairs$feature, sep="_") %in% unique(paste(errors$snp.chrom, errors$snp.pos, errors$restrict_gene, sep="_"))
missing_pairs$skip = paste(missing_pairs$ref_snp, missing_pairs$feature, sep="_") %in% unique(paste(skips$snp.chrom, skips$snp.pos, skips$feature, sep="_"))
sum(missing_pairs$error)
sum(missing_pairs$skip)
# Reasons why some combinations of trait-tissue at SNP-gene combos in our test set were skipped
table(skips$error)
```

Based on the QC checks here, I'm confident that we're not missing any tests for unexplainable reasons.

After performing all of these checks, we're ready to go on to the main colocalization analysis.

```{r include=FALSE}
missing_pairs$reason_missing = sapply(1:dim(missing_pairs)[1], function(i)
       {
		snp = missing_pairs$ref_snp[i]
       		gene = missing_pairs$feature[i]
		sub = results[(results$ref_snp == snp) & (results$feature == gene),]
		
		code = 0
		if(length(unique(sub$eqtl_file)) < 4)
		{
			code = code + 1
		}
		if(length(unique(sub$gwas_trait)) < 9)
		{
			code = code + 2
		}
		return(code)

       })

write.csv(missing_pairs[c("ref_snp", "feature", "reason_missing")], file="/users/mgloud/projects/insulin_resistance/output/all_missing_pairs_explained.txt", quote=FALSE, sep="\t", col.names=TRUE, row.names=FALSE)
write.csv(results, file="/users/mgloud/projects/insulin_resistance/output/full_coloc_results_qced.txt", quote=FALSE, sep="\t", col.names=TRUE, row.names=FALSE)
```
